Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-26 manuscript/paper.tex
first draft done. will share with KSabeeh and see if he likes something to change.

If we have an arrangement of ions with charge $q$ around a given site,
we can express their %(crystal field) 
potential in terms of spherical harmonics~\cite{PavariniChap}, 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
where $a$ sets the length scale
and $V_{l,m}$ are multipole components of the potential
 (for the given site as origin and a chosen orientation).
%and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we have a transition metal atom/ion
at the central site
with
only $d$-orbitals ($l=2$),
the matrix elements of its crystal field Hamiltonian
are then given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$
is the expectation of $(r/a)^l$ evaluated in the $d$-orbitals bases
assuming the same radial dependence for all orbitals~\cite{PaxtonNotes}.
The sum is restricted to even values of $l\leq 4$ as 
$C_{l}^{m'm''m}=0$ otherwise,
with $-l \leq m \leq l$.

@@

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six oxygen (O) ions each with a charge $-q_o$ making an octahedron of size $a$ (where these anions are placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the $d$-orbitals contains 
 only two $l=4$ components, 
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$),
 and is called octahedral field. 
It is given by~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds: 
% $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

Let us consider elongation of an octahedron % in sec.~\ref{sec:egt2g} 
along one of its O-B-O bonds. 
Align the cartesian axes along O-B-O bonds and stretch the two bonds along $z$-axis equally by an amount $a\delta$.
%where $a$ is the size of the octahedron (the distance between the O-O along any cartesian axes).
The potential $V^{JT}(\vec r)$ at the central B site (JT ion) of this deformed octahedron 
still contains only a few non-zero components.
Not only the ratio 
$V_{4,0}^{JT}:V_{4,4}^{JT}$ now deviates from
 ideal 
octahedral field in Eq.~\ref{eq:voct},
there is an $l=2$ component induced due to the deformation,
\begin{align}
%V_{0,0}^{JT}=& \frac{8 \sqrt{\pi } (2 \delta +3)}{\delta +1},\\
V_{2,0}^{JT}=& -\frac{32 \sqrt{\frac{\pi }{5}} \delta  \left(\delta ^2+3 \delta +3\right)}{(\delta +1)^3},\\
V_{4,0}^{JT}=& \frac{32}{3} \sqrt{\pi } \left(\frac{4}{(\delta +1)^5}+3\right),\\
V_{4,4}^{JT}=& \frac{32 \sqrt{35 \pi }}{3}.
\end{align}

@@

The crystal field Hamiltonian of the JT ion,
${H_{m',m''}^{JT} = (q_o/a)\sum_{l,m} V_{l,m}^{JT} C_{l}^{m'm''m} D_{l}}$,
still turns out to be diagonal 
(i.e., the deformation $\delta$
will not couple these orbitals to each other),
but the degeneracies of the two manifolds is lifted
and their respective average energies also shifted.
Defining 
$F_n(\delta) = 1-1/(1+\delta)^n$,
we see that
the $t_{2g}$ now splits into two levels,
with energies
\begin{align}
E_{xy} =&
\frac{32}{7}\frac{q_o}{a} \left(
 D_2F_3 - \frac{20}{9}D_4F_5 
\right),\\
E_{yz} =& E_{zx} = -\frac{1}{2} E_{xy},
\end{align}
as measured from their average 
$E_0^{t_{2g}} = -64q_oD_4 \left(3-F_5\right)/9a$.
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&
\frac{32}{7}\frac{q_o}{a} \left(D_2F_3+\frac{5}{3} D_4F_5\right),\\
E_{z^2}=&-E_{x^2-y^2},\\
\end{align}
as measured from their average 
$E_0^{e_{g}} = 32 q_oD_4 \left(3-F_5\right)/3a$.
The crystal field splitting becomes
$\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=
160q_oD_4\left(3-F_5\right)/9a$.
At small $\delta$, 
expanding 
$F_3,F_5$
 up to first order in $\delta$,
we obtain
\begin{align}
\label{eq:JTt2g}
(E_{xy},E_{yz},E_{zx})&= (2, -1, -1)
%\left[\frac{16}{63} (27 \text{D2}-100 \text{D4})\right]
\left[D_2-\frac{100}{27}  D_4\right]\frac{48}{7}\frac{q_o}{a}\delta,\\
\label{eq:JTeg}
(E_{x^2-y^2},E_{z^2})&=(1,-1)
\left[D_2+\frac{25}{9} D_4\right]\frac{96}{7}\frac{q_o}{a}\delta,
\end{align}
whereas
%${\Delta_{\delta}= \frac{48}{7}\frac{q_o}{a}\frac{\delta}{a}}$.
$E_0^{t_{2g}}$, $E_0^{e_{g}}$ and $\Delta$ all are rescaled by a factor of ${1-5\delta/3}$.

@@

Including the contribution of these AB cages to the crystal field,
we obtain
\begin{align}
%V(\vec r) &= V_{O}(\vec r) + V_{AB}(\vec r),\\
V(\vec r)  &= (1+f_{q_A}) V_{O}(\vec r),
\end{align}
whereas the 
 splitting becomes
%\begin{align}
$\Delta = (1+f_{q_A})\Delta_o$.
%\end{align}
%where ${\Delta_O=\frac{160}{3}\frac{q_o}{a} D_{4}}$.
We see that, relative to the O octahedron,
 the contribution of the
 two cation cages towards the total field splitting
 is relatively small at any $q_A$ but
 still large enough to be ignored.
It can be measured by $(\Delta-\Delta_o)/\Delta_o=f_{q_A}$,
which linearly changes from 
$-\frac{3}{32}\simeq -10\%$ at $q_A=0$
to $\frac{8}{27 \sqrt{3}}\simeq 17\%$ at $q_A=3q_o$. 
Typical values of $q_A$ relevant for real transition metal oxides
can be safely assumed to range between $0$ and $4$;  
$q_A=0$ for systems like tungsten oxide $WO_3$ that do not have any cations at the A sites,
and $q_A=4$ still contributing just above $8\%$.
We will see that,
as the anion octahedra rotate or tilt,
$V_{AB}(\vec r)$ rotate in its frame and deform as well.
The deformation also induces $l=2$ components, which couple to the 
$d$-orbitals of the B ion much more strongly. 
%due to a much larger $D_{l=2}$.

@@

It is worth reminding that
$V(\vec r) $ a scalar quantity in the position space but a vector 
%$V$ is as a vector %of length $(l_{max}+1)^2=25$ for $l_{max}=4$.
in the spherical harmonics bases, so
we can calculate it in one bases as per our convenience and then transform it to another.
%Its components in the new basis.
\begin{align}
V(\vec r) &= \sum_{l,m} V_{lm} r^l Y_{lm}(\hat r)
= \sum_{l,m} \tilde V_{lm} r^l \tilde Y_{lm}(\hat r),\\
%\tilde V_{lm} &= \sum_{l',m'}\mathcal{R}_{lm,l'm'}V_{l'm'},
\tilde V_{lm} &= \sum_{m'=-l}^{l}\mathcal{R}^{l}_{m,m'}V_{lm'},
\end{align}
where 
$\mathcal{R}^{l}=\mathcal{R}^{l}(\alpha,\beta)$ is the rotation matrix for real spherical harmonics 
of degree $l$. % (see appendix~\ref{sec:appendix}).
It can be calculated using the method 
devised by Ivanic and Ruedenberg~\cite{IvanicJPC96}.
The multipole components of $V_O(\vec r)$ in the rotated frame do not change with the rigid octahedral rotations because, by definition, the octahedron itself rotates with its frame.
These can be directly read off Eq.~\ref{eq:voct}.
Similarly, it is easier to obtain the components of $V_{AB}(\vec r)$
in the unrotated frame so we do that first and later transforms them to obtain the rotated frame components.
That is,
\begin{gather}
\tilde V_{lm} = \tilde V_{lm}^O + \sum_{m'=-l}^{l}\mathcal{R}^{l}_{m,m'} V_{lm'}^{AB},\\
V_{lm}^{AB} = \left(\frac{q_A}{a} \right)V_{lm}^{A}   
+ \left(\frac{q_B}{a}\right)V_{lm}^{B},
\end{gather}
%where $q_A + q_B = 3q_o$ ensures charge neutrality.
where the quantities bearing a tilde on top belong to the rotated (octahedron's) frame.
As described above, 
$\tilde V_{lm}^O$ are simply octahedral field components 
with $l=4,m=0,4$ (see Eq.~\ref{eq:voct}).
However, $V_{lm}^{AB}$ contain the effect of the deformation of the crystal structure 
%on AB ion cages due to the octahedral rotations
so it now has not only other $l=4$ components but also $l=2$ components; see below.
A procedure to obtain these multipole components, without
actually calculating the projection of the potential $V_{AB}(\vec r)$ 
onto the spherical harmonics, is described in the appendix. %~\ref{append:multipoles}.

@@

The Hamiltonian of the central ion (JT ion) including 
the crystal field of the AB cages 
in the octahedron's frame %(rotated frame)
are given by,
\begin{align}
\tilde H_{m',m''} &= \sum_{l,m} \tilde V_{l,m} \tilde C_{l}^{m'm''m} D_{l} ,
%C_{l}^{m'm''m} &= \braket{\tilde Y_{2,m'}|\tilde Y_{l,m}|\tilde Y_{2,m''}},
\end{align}
where %$C_{l}^{m'm''m}$ 
$\tilde C_{l}^{m'm''m}=\braket{\tilde Y_{2,m'}|\tilde Y_{l,m}|\tilde Y_{2,m''}}= C_{l}^{m'm''m}$. 
% are again the Gaunt coefficients for the real spherical harmonics.

@@

Before we present our results,
let us briefly describe 
the kind of behaviour that would indicate a JT instability 
against the tilting and rotations of the octahedra,
and how it relates to the standard JT problem.
As describe in sec.~\ref{sec:JT},
the optimum structure can have a finite distortion
as long as 
the splitting in the single particle spectrum has a different dependence on the distortion
than the elastic energy associated with the changes in the unit cell volume.
In contrast to 
%hypothetical 
elongations or compressions 
$a\delta_1,a\delta_2$ 
of an octahedron along its axes,
where the 
 the unit cell volume changes
as $\sim a^3\left(\delta_1+\delta_1\right)$
and thus the elastic energy approximately goes
as $\sim\left(\delta_1+\delta_1\right)^2$,
the unit cell volume,
in case of octahedral rotation and tilting,
 depends on the ``distortions'' $\alpha,\beta$ 
as ${v=4a^3 \cos ^2\alpha \cos ^2\beta}$,
and, at $\alpha,\beta\ll1$,
it changes as $\sim a^3\left(\alpha ^2+\beta ^2\right)$
leading to a quartic elastic energy that should roughly be
$\sim\left(\alpha ^2+\beta ^2\right)^2$.
So, the behaviour we are looking for in the electronic spectrum 
is not linear
but quadratic, 
which would compete with such a quartic elastic energy
 to produce a finite distortion in the optimum structure.
Thinking of squared angles as the basic variables,
the order of the electronic and elastic energy terms
matches to the standard JT problem.

@@

We will first focus on the $t_{2g}$ manifold
and see how the potential of the AB cages split it when 
the octahedra tilts and rotates.

@@

Ignoring the coupling between the $e_{g}$ and $t_{2g}$ manifolds (between $xy$ and $x^2-y^2$ at $\alpha=0$),
their Hamiltonians turn out to be diagonal.
We can write the single particle energies of $t_{2g}$ states
as
\begin{align}
\label{eq:theta0-Es}
(E_{xy},&E_{yz},E_{zx})= (2, -1, -1)
\left[\frac{}{} \Delta_{2}  D_2 + \Delta_4 D_4\right],
 \end{align}
 where $E_{yz}$ and $E_{zx}$ are degenerate and 
\begin{align}
\Delta_{l} =&  \left(q_A \Delta_{l}^A + q_B\Delta_{l}^B\right)/a,~(l=2,4),\\
\Delta_{2}^A =& 
%from A
\frac{64 \sin ^2\beta }{7 (\cos 2 \beta +2)^{5/2}} % simplifies.... from:
 %\frac{16 (3 \sin \beta+\sin 3\beta)^2}{7 (\cos 2\beta+2)^{9/2}},\\
 \simeq \frac{64}{63 \sqrt{3}}\beta ^2,\\
 \nonumber
\Delta_4^A =&
 \frac{160 \sin ^2\beta}{63 (\cos 2\beta+2)^{9/2}} \times \\
&  (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82)\\
\simeq& \frac{40960}{5103 \sqrt{3}}\beta ^2,\\
% \Delta_4^A =&
% \frac{160 \sin ^2\beta (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82)}{63 (\cos 2\beta+2)^{9/2}},\\
 %from B
 \label{eq:theta0-delB2}
\Delta_{2}^B =& \frac{2}{7} \left(1-\sec ^3\beta\right)
\simeq -\frac{3}{7}\beta ^2,\\
\label{eq:theta0-delB4}
  \Delta_4^B =&\frac{5}{126} \left((7 \cos 4\beta-3) \sec ^5\beta-4\right)
  \simeq -\frac{115}{63}\beta ^2.
 %\left( \Delta_{2}  D_2 + \Delta_4 D_4\right),\\
 \end{align}
 We see that $\Delta_{2/4}^A>0$ while $\Delta_{2/4}^B<0$,
 revealing a competition between the two terms 
 that would determine the sign and size of the term
  in the square bracket in Eq.~\ref{eq:theta0-Es} 
%  (hence the ground state and splitting) 
  as $q_A$ (and hence $q_B$)
  is varied.
The average $t_{2g}$ energy
 also shifts from
$-2/5\Delta$ by an amount
%\rev{E0t2g or change in it? how to define the expressions... plotted is always change in it...}
\begin{align}
\label{eq:theta0-E0}
E_{0}^{t_{2g}} = \left(q_A E_{0}^A + q_BE_{0}^B\right)D_4/a
-(2/5)f_{q_A}\Delta_{O},
 \end{align}
 where
\begin{align}
\nonumber
E_{0}^A =&
 \frac{8 (92 \cos 2\beta+24 \cos 4\beta+20 \cos 6\beta+5 \cos 8\beta+51)}{9 (\cos 2\beta+2)^{9/2}}\\
 \simeq& \frac{512}{243 \sqrt{3}}-\frac{2560 \beta ^2}{729 \sqrt{3}},\\
 \nonumber
E_{0}^B=&
 -\frac{1}{18} \left((5 \cos 4\beta+3) \sec ^5\beta+4\right)\\
 \simeq& -\frac{2}{3}+\frac{10 \beta ^2}{9}.
 \end{align}

>>>>

If we have an arrangement of ions with charge $q$ around a given site,
we can express their %(crystal field) 
potential in terms of spherical harmonics~\cite{PavariniChap}, 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{lm}~\left(\frac{r}{a}\right)^l \ylm{lm},
\end{align}
where $a$ sets the length scale
and $V_{l,m}$ are multipole components of the potential
 (for the given site as origin and a chosen orientation).
%and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we have a transition metal atom/ion
at the central site
with
only $d$-orbitals ($l=2$),
the matrix elements of its crystal field Hamiltonian
are then given by,
\begin{align}
H_{m' m''} &= -\frac{q}{a}\sum_{l,m} V_{lm} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2m'}| Y_{lm}| Y_{2m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$
is the expectation of $(r/a)^l$ evaluated in the $d$-orbitals bases
assuming the same radial dependence for all orbitals~\cite{PaxtonNotes}.
The sum is restricted to even values of $l\leq 4$ as 
$C_{l}^{m'm''m}=0$ otherwise,
with $-l \leq m \leq l$.

@@

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six oxygen (O) ions each with a charge $-q_o$ making an octahedron of size $a$ (where these anions are placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the $d$-orbitals contains 
 only two $l=4$ components, 
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$),
 and is called octahedral field. 
It is given by~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds: 
% $\Delta$ apart:
$H_{mm} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{mm} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

Let us consider elongation of an octahedron % in sec.~\ref{sec:egt2g} 
along one of its O-B-O bonds. 
Align the cartesian axes along O-B-O bonds and stretch the two bonds along $z$-axis equally by an amount $a\delta$.
%where $a$ is the size of the octahedron (the distance between the O-O along any cartesian axes).
The potential $V^{JT}(\vec r)$ at the central B site (JT ion) of this deformed octahedron 
still contains only a few non-zero components (see appendix for a method to calculate $V_{lm}^{JT}$).
Not only the ratio 
$V_{4,0}^{JT}:V_{4,4}^{JT}$ now deviates from
 ideal 
octahedral field in Eq.~\ref{eq:voct},
there is an $l=2$ component induced due to the deformation,
\begin{align}
%V_{0,0}^{JT}=& \frac{8 \sqrt{\pi } (2 \delta +3)}{\delta +1},\\
V_{2,0}^{JT}=& -\frac{32 \sqrt{\frac{\pi }{5}} \delta  \left(\delta ^2+3 \delta +3\right)}{(\delta +1)^3},\\
V_{4,0}^{JT}=& \frac{32}{3} \sqrt{\pi } \left(\frac{4}{(\delta +1)^5}+3\right),\\
V_{4,4}^{JT}=& \frac{32 \sqrt{35 \pi }}{3}.
\end{align}

@@

The crystal field Hamiltonian of the JT ion,
${H_{m' m''}^{JT} = (q_o/a)\sum_{l,m} V_{lm}^{JT} C_{l}^{m'm''m} D_{l}}$,
still turns out to be diagonal 
(i.e., the deformation $\delta$
will not couple these orbitals to each other),
but the degeneracies of the two manifolds is lifted
and their respective average energies also shifted.
Defining 
$F_n(\delta) = 1-1/(1+\delta)^n$,
we see that
the $t_{2g}$ now splits into two levels,
with energies
\begin{align}
E_{xy} =&
\frac{32}{7}\frac{q_o}{a} \left(
 D_2F_3 - \frac{20}{9}D_4F_5 
\right),\\
E_{yz} =& E_{zx} = -\frac{1}{2} E_{xy},
\end{align}
as measured from their average 
$E_0^{t_{2g}} = -64q_oD_4 \left(3-F_5\right)/9a$.
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&
\frac{32}{7}\frac{q_o}{a} \left(D_2F_3+\frac{5}{3} D_4F_5\right),\\
E_{z^2}=&-E_{x^2-y^2},\\
\end{align}
as measured from their average 
$E_0^{e_{g}} = 32 q_oD_4 \left(3-F_5\right)/3a$.
The crystal field splitting becomes
$\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=
160q_oD_4\left(3-F_5\right)/9a$.
At small $\delta$, 
expanding 
$F_3,F_5$
 up to first order in $\delta$,
we obtain
\begin{align}
\label{eq:JTt2g}
(E_{xy},E_{yz},E_{zx})&= (2, -1, -1)
%\left[\frac{16}{63} (27 \text{D2}-100 \text{D4})\right]
\left[D_2-\frac{100}{27}  D_4\right]\frac{48}{7}\frac{q_o}{a}\delta,\\
\label{eq:JTeg}
(E_{x^2-y^2},E_{z^2})&=(1,-1)
\left[D_2+\frac{25}{9} D_4\right]\frac{96}{7}\frac{q_o}{a}\delta,
\end{align}
whereas
%${\Delta_{\delta}= \frac{48}{7}\frac{q_o}{a}\frac{\delta}{a}}$.
$E_0^{t_{2g}}$, $E_0^{e_{g}}$ and $\Delta$ all are rescaled by a factor of ${1-5\delta/3}$.

@@

Including the contribution of these AB cages to the crystal field,
we obtain
\begin{align}
%V(\vec r) &= V_{O}(\vec r) + V_{AB}(\vec r),\\
V(\vec r)  &= (1+f_{q_A}) V_{O}(\vec r),
\end{align}
whereas the 
 splitting becomes
%\begin{align}
$\Delta = (1+f_{q_A})\Delta_o$.
%\end{align}
%where ${\Delta_O=\frac{160}{3}\frac{q_o}{a} D_{4}}$.
We see that, relative to the O octahedron,
 the contribution of the
 two cation cages towards the total field splitting
 is relatively small at any $q_A$ but
 still large enough to be ignored.
It can be measured by $(\Delta-\Delta_o)/\Delta_o=f_{q_A}$,
which linearly changes from 
$-3/32\simeq -10\%$ at $q_A=0$
to $8/{27 \sqrt{3}}\simeq 17\%$ at $q_A=3q_o$. 
Typical values of $q_A$ relevant for real transition metal oxides
can be safely assumed to range between $0$ and $4$;  
$q_A=0$ for systems like tungsten oxide $WO_3$ that do not have any cations at the A sites,
and $q_A=4$ still contributing just above $8\%$.
We will see that,
as the anion octahedra rotate or tilt,
$V_{AB}(\vec r)$ rotate in its frame and deform as well.
The deformation also induces $l=2$ components, which couple to the 
$d$-orbitals of the B ion much more strongly. 
%due to a much larger $D_{l=2}$.

@@

It is worth reminding that
$V(\vec r) $ a scalar quantity in the position space but a vector 
%$V$ is as a vector %of length $(l_{max}+1)^2=25$ for $l_{max}=4$.
in the spherical harmonics bases, so
we can calculate it in one bases as per our convenience and then transform it to another.
%Its components in the new basis.
\begin{align}
V(\vec r) &= \sum_{l,m} V_{lm} r^l Y_{lm}(\hat r)
= \sum_{l,m} \tilde V_{lm} r^l \tilde Y_{lm}(\hat r),\\
%\tilde V_{lm} &= \sum_{l',m'}\mathcal{R}_{lm,l'm'}V_{l'm'},
\tilde V_{lm} &= \sum_{m'=-l}^{l}\mathcal{R}^{l}_{m,m'}V_{lm'},
\end{align}
where 
$\mathcal{R}^{l}=\mathcal{R}^{l}(\alpha,\beta)$ is the rotation matrix for real spherical harmonics 
of degree $l$. % (see appendix~\ref{sec:appendix}).
It can be calculated using the method 
devised by Ivanic and Ruedenberg~\cite{IvanicJPC96}.
The multipole components of $V_O(\vec r)$ in the rotated frame do not change with the rigid octahedral rotations because, by definition, the octahedron itself rotates with its frame.
These can be directly read off Eq.~\ref{eq:voct}.
Similarly, it is easier to obtain the components of $V_{AB}(\vec r)$
in the unrotated frame so we do that first and later transforms them to obtain the rotated frame components.
That is,
\begin{gather}
\tilde V_{lm} = \tilde V_{lm}^O + \sum_{m'=-l}^{l}\mathcal{R}^{l}_{m m'} V_{lm'}^{AB},\\
V_{lm}^{AB} = \left(\frac{q_A}{a} \right)V_{lm}^{A}   
+ \left(\frac{q_B}{a}\right)V_{lm}^{B},
\end{gather}
%where $q_A + q_B = 3q_o$ ensures charge neutrality.
where the quantities bearing a tilde on top belong to the rotated (octahedron's) frame.
As described above, 
$\tilde V_{lm}^O$ are simply octahedral field components 
with $l=4,m=0,4$ (see Eq.~\ref{eq:voct}).
However, $V_{lm}^{AB}$ contain the effect of the deformation of the crystal structure 
%on AB ion cages due to the octahedral rotations
so it now has not only other $l=4$ components but also $l=2$ components; see below.
A procedure to obtain these multipole components, without
actually calculating the projection of the potential $V_{AB}(\vec r)$ 
onto the spherical harmonics, is described in the appendix. %~\ref{append:multipoles}.

@@

The Hamiltonian of the central ion (JT ion) including 
the crystal field of the AB cages 
in the octahedron's frame %(rotated frame)
are given by,
\begin{align}
\tilde H_{m' m''} &= \sum_{l,m} \tilde V_{l m} \tilde C_{l}^{m'm''m} D_{l} ,
%C_{l}^{m'm''m} &= \braket{\tilde Y_{2,m'}|\tilde Y_{l,m}|\tilde Y_{2,m''}},
\end{align}
where %$C_{l}^{m'm''m}$ 
$\tilde C_{l}^{m'm''m}=\braket{\tilde Y_{2m'}|\tilde Y_{lm}|\tilde Y_{2m''}}= C_{l}^{m'm''m}$. 
% are again the Gaunt coefficients for the real spherical harmonics.

@@

Before we present our results,
let us briefly describe 
the kind of behaviour that would indicate a JT instability 
against the tilting and rotations of the octahedra,
and how it relates to the standard JT problem.
As describe in sec.~\ref{sec:JT},
the optimum structure can have a finite distortion
as long as 
the splitting in the single particle spectrum has a different dependence on the distortion
than the elastic energy associated with the changes in the unit cell volume.
In contrast to 
%hypothetical 
elongations or 
compressions---$a\delta_1,a\delta_2$ 
of an octahedron along two of its axes---where the 
 the unit cell volume changes
as $\sim a^3\left(\delta_1+\delta_1\right)$
and thus the elastic energy approximately goes
as $\sim\left(\delta_1+\delta_1\right)^2$,
the unit cell volume
in case of octahedral rotation and tilting
 depends on the ``distortions'' $\alpha,\beta$ 
as ${v=4a^3 \cos ^2\alpha \cos ^2\beta}$,
and, at $\alpha,\beta\ll1$,
it changes as $\sim a^3\left(\alpha ^2+\beta ^2\right)$
leading to a \emph{quartic} elastic energy that should roughly behave as
$\sim\left(\alpha ^2+\beta ^2\right)^2$.
So, the behaviour we are looking for in the electronic spectrum 
is not linear
but quadratic, 
which would compete with such a quartic elastic energy
 to produce a finite distortion in the optimum structure.
Thinking of squared angles 
$\alpha ^2,\beta ^2$
as the basic distortion variables,
the order of the electronic and elastic energy terms
matches to the standard JT problem.

@@

We will first focus on the $t_{2g}$ manifold
and see how the potential of the AB cages split it when 
the octahedra tilt and rotate.

@@

Ignoring the coupling between the $e_{g}$ and $t_{2g}$ manifolds (between $xy$ and $x^2-y^2$ at $\alpha=0$),
their Hamiltonians turn out to be diagonal.
We can write the single particle energies of $t_{2g}$ states
as
\begin{align}
\label{eq:theta0-Es}
(E_{xy},&E_{yz},E_{zx})= (2, -1, -1)
\left[\frac{}{} \Delta_{2}  D_2 + \Delta_4 D_4\right],
 \end{align}
 where $E_{yz}$ and $E_{zx}$ are degenerate and 
\begin{align}
\Delta_{l} =&  \left(q_A \Delta_{l}^A + q_B\Delta_{l}^B\right)/a,~(l=2,4),\\
\Delta_{2}^A =& 
%from A
\frac{64 \sin ^2\beta }{7 (\cos 2 \beta +2)^{5/2}} % simplifies.... from:
 %\frac{16 (3 \sin \beta+\sin 3\beta)^2}{7 (\cos 2\beta+2)^{9/2}},\\
 \simeq \frac{64}{63 \sqrt{3}}\beta ^2,\\
 \nonumber
\Delta_4^A =&
 \frac{160 \sin ^2\beta}{63 (\cos 2\beta+2)^{9/2}} \times \\
&  (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82)\\
\simeq& \frac{40960}{5103 \sqrt{3}}\beta ^2,\\
% \Delta_4^A =&
% \frac{160 \sin ^2\beta (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82)}{63 (\cos 2\beta+2)^{9/2}},\\
 %from B
 \label{eq:theta0-delB2}
\Delta_{2}^B =& \frac{2}{7} \left(1-\sec ^3\beta\right)
\simeq -\frac{3}{7}\beta ^2,\\
\label{eq:theta0-delB4}
  \Delta_4^B =&\frac{5}{126} \left((7 \cos 4\beta-3) \sec ^5\beta-4\right)
  \simeq -\frac{115}{63}\beta ^2.
 %\left( \Delta_{2}  D_2 + \Delta_4 D_4\right),\\
 \end{align}
 We see that $\Delta_{2/4}^A>0$ while $\Delta_{2/4}^B<0$,
 revealing a competition between the two terms 
 that would determine the sign and size of the term
  in the square bracket in Eq.~\ref{eq:theta0-Es} 
%  (hence the ground state and splitting) 
  as $q_A$ (and hence $q_B$)
  is varied.
The average $t_{2g}$ energy
 also shifts from
$-2/5\Delta$ by an amount
%\rev{E0t2g or change in it? how to define the expressions... plotted is always change in it...}
\begin{align}
\label{eq:theta0-E0}
E_{0}^{t_{2g}} = \left(q_A E_{0}^A + q_BE_{0}^B\right)D_4/a
-(2/5)f_{q_A}\Delta_{o},
 \end{align}
 where
\begin{align}
\nonumber
E_{0}^A =&
 \frac{8 (92 \cos 2\beta+24 \cos 4\beta+20 \cos 6\beta+5 \cos 8\beta+51)}{9 (\cos 2\beta+2)^{9/2}}\\
 \simeq& \frac{512}{243 \sqrt{3}}-\frac{2560 \beta ^2}{729 \sqrt{3}},\\
 \nonumber
E_{0}^B=&
 -\frac{1}{18} \left((5 \cos 4\beta+3) \sec ^5\beta+4\right)\\
 \simeq& -\frac{2}{3}+\frac{10 \beta ^2}{9}.
 \end{align}

The results in Eqs.~\ref{eq:theta0-Es},\ref{eq:theta0-E0}
are summarised in Fig.~\ref{fig:t2g-alpha0},
where $E_{xy}$, $E_{yz/zx}$ and $E_0^{t_{2g}}$ are shown
as a function of $\beta$
at three values of $q_A/q_o=1/2,1,2$.
We have 
assumed $D_2=\sqrt{D_4}$ and
taken $\Delta_o=4q_o/15a$ throughout this paper,
which fixes the value of $D_4=1/200$, 
and corresponds to $\Delta_o= 2.073$ eV at 
$q_o=2$ e and $a=7.0$ bohr.
We present the energies as a percentage of $\Delta_o$
by calculating all quantities at the above values of $q_o,a$.
Fig.~\ref{fig:t2g-alpha0} shows that
the competition between A and B cages
determines the ground state and the size of the splitting between the 
$t_{2g}$ states
$E_{xy}$ and $E_{yz/zx}$.
In Fig.~\ref{fig:t2g-alpha0}(a),
where $q_A/q_o=1/2$ ($q_B/q_o=5/2$),
the effect of B cage dominates.
Since it's negative (see Eqs.~\ref{eq:theta0-delB2},\ref{eq:theta0-delB4}),
it reverses the order in Eq.~\ref{eq:theta0-Es}
and $E_{xy}$ becomes lower than $E_{yz/zx}$, by about $6\%$ of $\Delta_o$ at $\beta=15^\circ$.
When A cage dominates at large $q_A/q_o$,
as shown in Fig.~\ref{fig:t2g-alpha0}(c) for $q_A/q_o=2$,
$E_{yz/zx}$ are the single particle ground states
 (with a splitting of the same order), %$\sim6\%$ of $\Delta_o$ at $15^\circ$),
whereas the net effect of both cages
is reduced in Fig.~\ref{fig:t2g-alpha0}(b)
that corresponds to $q_A/q_o=1$
and we only obtain a splitting of $\sim2\%$ of $\Delta_o$ at $\beta=15^\circ$.
%between $E_{xy}$ and $E_{yz/zx}$ levels.
Obviously, there will also be a point near 
$q_A/q_o=1$ where A and B completely 
cancel the effect of each other (not shown in Fig.~\ref{fig:t2g-alpha0}).
We also see that $E_0^{t_{2g}}$ changes from a small negative to a small positive value
with an increase in $q_A/q_o$ as we
go from Fig.~\ref{fig:t2g-alpha0}(a)
to Fig.~\ref{fig:t2g-alpha0}(c).
This is because the rotation suppresses the
effect of AB cages on the
gap $\Delta$ between $e_{g}$ and $t_{2g}$.
Since A supports O while B opposes it,
$\Delta$ slightly widens at small $q_A$ and shrinks at large $q_A$,
shifting the $t_{2g}$ manifold slightly down or up with $\beta$.

These results also indicate that in case of a superlattice with $q_A$ alternating
between the two regimes shown in Figs.~\ref{fig:t2g-alpha0}(a,c),
we should obtain an alternating orbital order if the octahedra are rotated.
